Community context determines competition/facilitation trade-offs in pollination systems

### 2019-04-28 paper/manuscript.Rmd
dbs edits and comments prior to the fire alarm

Animal pollination systems play a disproportionally important role in food production and maintenance of global biodiversity [@bascompte_plant-animal_2007; @klein_importance_2007; @ollerton_how_2011]. 
<!-- FDO: Asked coauthors for a reference for the following statement but no-one provided. Therefore removing -->
Alongside the direct mutualisms between plants and their pollinators, biodiversity can be supported as a result of the inter-species "facilitation"—the indirect positive feedback loops that exist between plant species that share pollinators or pollinators that share plants [@moeller_facilitative_2004; @ghazoul_floral_2006; @molina-montenegro_positive_2008; @liao_presence_2011]. 
This inter-species facilitation is able to promote species coexistence by offsetting the effects of direct resource competition [@stachowicz_mutualism_2001]. 
Indeed, studies at the core of our current theoretical knowledge of mutualisms predict that the maximum number of coexisting species is achieved when the number of shared mutualistic partners is maximised in a community [@bastolla_architecture_2009].

However, in natural communities, plants rarely share all available pollinators.
One explanation is that plants are effectively competing for mutualistic partners because there is a *trade-off* between the benefits gained from maximising the number of partners and the costs of sharing them with other plant species [@waser_interspecific_1978]. 
It has been clear for a long time that when competition for pollination is strong enough, these costs can make it detrimental to plants' fitness [shown both experimentally and theoretically; @robertson_philosophy_1895; @lewis_experimental_1961; @levin_competition_1970]. 
In fact, it can be sufficient to drive ecological differentiation [for example by driving the evolution of both traits and reproductive strategies; @mitchell_new_2009; @caruso_competition_2000]. 
It is less clear, however, how exactly ecological differentiation, in turn, affect the trade-off between facilitation and competition that is involved in the pollination service.

There are two main mechanisms through which competition for pollinators can affect plant fertilization [@morales_interspecific_2008]. 
The first is intraspecific pollen transfer. 
This happens, for example, when plants with distinctively attractive flowers might reduce the number of visits to those less attractive neighbouring plants, and hence reduce the amount of *conspecific pollen* deposited by animals [@yang_pollinator_2011]. 
The second is intraspecific pollen transfer and occurs when plants share pollination partners. 
In that case, even receiving a visit might not necessarily translate into fertilization [@campbell_mechanism_1985] because a focal plant might receive *heterospecific pollen* or because pollen from the focal plant might be lost to different species. 
Generally speaking, the higher the amount of conspecific pollen (both in absolute terms and relative to heterospecific pollen) the better the pollination service received by the focal plant.
These two mechanisms of competition, by definition, occur at the community scale. 
However, with few exceptions [@rathcke_interactions_1988; @lopezaraiza-mikel_impact_2007; @tur_evaluating_2016; @hegland_relative_2009; @aizen_reproductive_2010], most of what we know about the deposition of conspecific and heterospecific pollen---and its relationship to competition vs. facilitation in pollination systems---is based on studies with two plant species.

At the community scale, the factors that determine the patterns of conspecific and heterospecific pollen deposition are tightly intertwined, operate simultaneously, and may lead to emergent phenomena not observed at smaller scales [@flanagan_effects_2011]. 
For instance, recent empirical evidence suggests that plants with flowering traits that are "original" relative to others in the community generally have fewer interaction partners [@coux_linking_2016]. 
This evidence is aligned with the notion that a species which interacts with few species does so strongly with each of them whereas a species that interacts with a large number of species does so comparatively weakly\R{ref-interaction-strength}. 
If evolutionary specialisation occurs by changing traits to focus on fewer but better partners [@caruso_competition_2000], we should expect a reduction of competition for pollinators in species with "original" traits and an increase of competition in species with a large number of interaction partners [@gibson_can_2012; @carvalheiro_potential_2014]. 
Alternatively, it might also be the case that density (for example in terms of flower or pollen counts) is the dominant force driving pollen transfer [@seifan_outcome_2014]. 
Abundant plant species might experience a dilution of available pollinators [@feinsinger_effects_1987] but might also receive more effective visits by capitalising on a larger share of both visits and the pollen carried by pollinators. 
In this case, a potential reduction in the absolute amount of conspecific pollen received could be compensated by an increase in the amount of conspecific pollen relative to heterospecific pollen. 
Altogether, it is clear that some ecological factors can influence pollen deposition and that these factors, in the specific context of pollination, could be used to roughly define a species *realised niche* in their community. 
However, the intertwined nature of these factors makes it hard to determine whether they favour or hinder the pollination service at the community level. 
Moreover, their relative importance has not yet been empirically evaluated in natural plant communities.

Here, we investigate competition for pollination at the community scale using empirical data from eleven plant-pollinator communities in the Argentinian Pampas. 
First, we estimate the extent to which the net outcome of plants sharing pollinators is facilitation or competition. 
If facilitation is the primary outcome, we could expect two things. 
On the one hand, we could expect an overall positive relationship between the amount of conspecific pollen deposited in stigmas and the amount of heterospecific pollen [@tur_evaluating_2016]. 
The higher the deposition of conspecific pollen relative to heterospecific, the larger the slope of that relationship.
On the other, we could also expect a larger amount of conspecific pollen deposited in stigmas when flowers are open to animal pollinators than what is deposited due to self-pollination. 
Second, we investigate the relative contribution that four ecological factors have on the pollination service. 
Specifically, we hypothesize that competition, measured as the quantity of conspecific pollen deposition and its purity (relative to heterospecific pollen), should increase for plants that share many pollination partners.
However, we also hypothesize that other factors like the plant's functional originality, its relative abundance, and the visit effectiveness should have the potential to compensate for this increase in competition. 
Third and finally, we examine how much does the community context influences the way plants use these factors to minimise competition. 
Although some plant species might be widespread across multiple locations, competition for pollinators occurs at the community level and hence we predict the competition for pollination niche to be influenced by the local context. 
If plant strategy is not flexible across communities\R{ref-plant-flexibility}, we could expect niches to be more similar within species than across species in the study. 
On the contrary, if a species' strategy is flexible, we should expect plants to be able to occupy different niches in each community such that competition could be minimised regardless of the context.

@@

We collected data from eleven co-flowering plant communities and their pollinators in three locations in the Argentinian Pampas. 
These locations are distributed across a precipitation gradient with a humid temperate climate of the eastern Flooding Pampas becoming drier as one moves inland and westward towards the Pampean Grasslands. 
This precipitation gradient generated a natural gradient of network architecture which allowed us to examine the generality of our findings. 
In each location, we sampled two restored and two agricultural fragments, except in the Flooding Pampas, where we were only able to sample one restored fragment due to the lack of available sites.

@@

Our first objective was to identify whether the outcome of plants sharing pollinators at the community level is facilitation or competition. 
We use two complementary lines of evidence.  
The first line can be found via the relationship between heterospecific and conspecific pollen deposited in plant stigmas [@tur_evaluating_2016]. 
This is because heterospecific pollen can be seen as an unavoidable cost that plants must pay for receiving conspecific pollen from shared pollinators. 
A negative relationship (when shared pollinators increase heterospecific pollen loads and decrease conspecific loads) may indicate that competition for pollination is strong.
In contrast, a positive relationship is what would be observed when facilitation is the dominant outcome.

@@

In each of the studied communities, we quantified pollen deposition in a subset of plant species between December 2010 and February 2011. 
This subset comprised between three and nine common insect-pollinated (entomophilous) plant species that were flowering during the sampling period (see Figure \@ref(fig:fig-trade-off-predictions)a). 
We chose plant species such that they cover a wide range on a specialization-generalization gradient as well as a wide range of abundances. 
In each of the selected plants, we removed all flowers leaving only buds that were expected to go into florescence on the next day. 
Approximately a quarter of these buds were bagged to prevent animal pollination. 
Two days after inflorescence, we collected all remaining flowers' and counted the number of conspecific and heterospecific pollen grains in their pistils. 
More details can be found in @marrero_agricultural_2016.

@@

To evaluate the relationship between heterospecific and conspecific pollen, we used a generalised linear mixed model (GLMM) with a Poisson error distribution [@tur_evaluating_2016]. 
In this model, the number of conspecific pollen grains per flower was the response variable and the number of heterospecific grains the predictor (the number of stigmas sampled was added as an offset term). 
The slope and intercept of this relationship were allowed to differ between plant species within each community, which were modelled as a random effect. 
Models were fitted using the function `glmer` from the R package `lme4 1.1-19` [@bates_fitting_2015]. 
A positive slope for a plant-community combination might indicate facilitation while a negative slope might indicate competition [@tur_evaluating_2016].

To compare the conspecific pollen deposited with and without animal-mediated pollination, we followed the same approach as above. 
In this case, however, the predictor was the treatment (whether the flower was bagged or unbagged) instead of the number of heterospecific grains.

@@

Our second objective was to investigate the relative contribution of different factors---that describe the plant competition for pollination niche---to the pollination service. 
Generally speaking, in the context of competition for pollination, we expect that a factor that increases the amount of conspecific pollen deposited in stigmas, both in quantity and purity relative to heterospecific pollen, also has a positive effect on the pollination service. 
Specifically, we investigated the effect of (*i*) a plant's number of shared pollinator species, (*ii*) a plants abundance relative to the rest of the community, (*iii*) the mean visit effectiveness---a metric that combines the share of pollen that a plant species is able to place on each of their floral visitors and the number of visits it receives from them, and (*iv*) the plants functional originality [@laliberte_distance-based_2010]. 
See *Data Analysis* section below for more details on these four variables. 
<!-- Here, we define visit effectiveness of the pollinator species *i* to plant species *j* as the proportion between the visits of *i* to *j* and all visits of *i*, multiplied by the proportion between the number of pollen grains of *j* attached to individuals of *i* and all pollen attached to individuals of *i*. -->

@@

To obtain the number of shared pollinators for each species, we collected data to construct qualitative and quantitative pollination networks. 
Qualitative networks were constructed based on two-hour observations of floral visits in each fragment. 
Quantitative networks were constructed using two 50 m randomly located transects in each fragment. 
We counted and collected all floral visitors found in a 2 m wide strip while walking at a pace of 10 m per minute [@marrero_effect_2014; @memmott_structure_1999]. 
We visited the transects each month between November 2010 and March 2011. 
To obtain plant abundance, we counted all units of floral attraction found during an independent sampling of the transects used to construct the quantitative visitation networks. 
To estimate visit effectiveness we need to construct pollen transfer networks in addition to the visitation networks. 
To do this, we examined the pollen loads present on the floral visitors collected [@marrero_exotic_2017]. 
When the pollen count on an individual animal was estimated to be less than 2,000 grains, we identified every grain to the species level when possible and to pollen complexes when it was not. 
When the pollen count was above 2,000 grains, we classified approx. 50% of pollen and total pollen counts were extrapolated [@bosch_plant-pollinator_2009]. 
Finally, we also recorded morphological traits that relate to plant type (herb, shrub, climber), life cycle (annual, perennial), flower colouration, phenology, and whether the species is native in the study region\R{justification-traits}. 
<!-- We used those traits because they were the only ones available. <!--.. (TODO: Add a better justification for the traits used)--> 
More details can be found in Marrero *et al.* (2014 and 2017).

@@

To investigate the impact of ecological factors on pollination service we used two sets of linear mixed models (LMM) with bootstrap resampling.
The response variables for these model sets were the number of conspecific and heterospecific pollen grains deposited per stigma in flowers open to animal-mediated pollination. 
We used LMMs in which pollen loads were log-transformed because these models offered a better fit than a GLMM with Poisson (or quasipoisson) error structure. 
Models were fitted using the R package `nlme 3.1-131` [@pinheiro_nlme_2018].

>>>>

Animal pollination plays a disproportionally important role in food production and maintenance of global biodiversity [@bascompte_plant-animal_2007; @klein_importance_2007; @ollerton_how_2011]. 
<!-- FDO: Asked coauthors for a reference for the following statement but no-one provided. Therefore removing -->
Alongside the direct benefits of mutualisms between plants and their pollinators, biodiversity can be supported as a result of the inter-species "facilitation"—the indirect positive feedback loops that exist between plant species that share pollinators or pollinators that share plants [@moeller_facilitative_2004; @ghazoul_floral_2006; @molina-montenegro_positive_2008; @liao_presence_2011]. 
This inter-species facilitation is able to promote species coexistence by offsetting the effects of direct resource competition [@stachowicz_mutualism_2001]. 
Indeed, studies at the core of our current theoretical knowledge of mutualisms predict that the maximum number of coexisting species is achieved when the number of shared mutualistic partners is maximised in a community [@bastolla_architecture_2009].

However, in natural communities, plants rarely share all available pollinators.
One explanation is that plants are effectively competing for mutualistic partners because there is a *trade-off* between the benefits gained from maximising the number of partners and the costs of sharing them with other plant species [@waser_interspecific_1978]. 
It has been clear for a long time that when competition for pollination is strong enough, these costs can make pollination detrimental to plants' fitness [shown both experimentally and theoretically; @robertson_philosophy_1895; @lewis_experimental_1961; @levin_competition_1970]. 
In fact, it can be sufficient to drive ecological differentiation [for example by driving the evolution of both traits and reproductive strategies; @mitchell_new_2009; @caruso_competition_2000]. 
However, it is less clear how exactly ecological differentiation, in turn, affects the trade-off between facilitation and competition that is involved in the pollination service.
<!-- DBS; The end of the sentence above is confusing. Is it because what you mean by "the pollination service" hasn't been made explicit yet? -->

There are two main mechanisms through which competition for pollinators can affect plant fertilization [@morales_interspecific_2008]. 
The first is by changes in intraspecific pollen transfer.
<!-- DBS: The above should be interspecific, no? -->
This happens, for example, when plants with more attractive flowers might reduce the number of visits to those less of attractive neighbouring plants, and hence reduce the amount of *conspecific pollen* deposited by animals [@yang_pollinator_2011]. 
The second is via intraspecific pollen transfer and occurs when plants share pollination partners. 
In that case, even receiving a visit might not necessarily translate into fertilization [@campbell_mechanism_1985] because a focal plant might receive *heterospecific pollen* or because pollen from the focal plant might be lost to different species. 
Generally speaking, the higher the amount of conspecific pollen (both in absolute terms and relative to heterospecific pollen), the better the pollination service received by the focal plant.
By definition, these two mechanisms of competition occur at the community scale. 
However, with few exceptions [@rathcke_interactions_1988; @lopezaraiza-mikel_impact_2007; @tur_evaluating_2016; @hegland_relative_2009; @aizen_reproductive_2010], most of what we know about the deposition of conspecific and heterospecific pollen---and its relationship to competition vs. facilitation in pollination communities---is based on studies with two plant species.

At the community scale, the factors that determine the patterns of conspecific and heterospecific pollen deposition are tightly intertwined, operate simultaneously, and may lead to emergent phenomena not observed at smaller scales [@flanagan_effects_2011].
<!-- DBS: Fer, I absolutely love this sentence. -->
For instance, recent empirical evidence suggests that plants with flowering traits that are "original" relative to others in the community generally have fewer interaction partners [@coux_linking_2016]. 
This evidence is aligned with the notion that a species that interacts with few species does so strongly with each of them whereas a species that interacts with a large number of species does so comparatively weakly\R{ref-interaction-strength}. 
<!-- DBS: Doesn't Bascompte et al 2006 Science show this? Or maybe Thébault and Fontaine 2008 Oikos? -->
If evolutionary specialisation occurs by changing traits to focus on fewer but better partners [@caruso_competition_2000], we should expect a reduction of competition for pollinators in species with "original" traits and an increase of competition in species with a large number of interaction partners [@gibson_can_2012; @carvalheiro_potential_2014]. 
Alternatively, it might also be the case that abundance (for example in terms of flower or pollen counts) is the dominant force driving pollen transfer [@seifan_outcome_2014]. 
Abundant plant species might experience a dilution of available pollinators [@feinsinger_effects_1987] but might also receive more effective visits by capitalising on a larger share of both visits and the pollen carried by pollinators. 
In this case, a potential reduction in the absolute amount of conspecific pollen received could be compensated by an increase in the amount of conspecific pollen relative to heterospecific pollen.
<!-- DBS: I feel like the paragraph starts to lose the plot in the bit that follows. Where did realised niche come from and why invoke it? -->
Altogether, it is clear that some ecological factors can influence pollen deposition and that these factors, in the specific context of pollination, could be used to roughly define a species *realised niche* in their community. 
However, the intertwined nature of these factors makes it hard to determine whether they favour or hinder the pollination service at the community level. 
Moreover, their relative importance has not yet been empirically evaluated in natural plant communities.

Here, we investigate competition for pollination at the community scale using empirical data from eleven plant-pollinator communities in the Argentinian Pampas. 
First, we estimate the extent to which the net outcome of plants sharing pollinators is facilitation or competition. 
If facilitation is the primary outcome, we could expect two things. 
On the one hand, we could expect an overall positive relationship between the amount of conspecific pollen deposited in stigmas and the amount of heterospecific pollen [@tur_evaluating_2016]. 
The higher the deposition of conspecific pollen relative to heterospecific, the larger the slope of this relationship.
On the other hand, we could expect a larger amount of conspecific pollen deposited in stigmas when flowers are open to animal pollinators than what is deposited due to self-pollination. 
Second, we investigate the relative contribution that four ecological factors have to the pollination service. 
Specifically, we hypothesize that competition, measured as the quantity of conspecific pollen deposition and its purity (amount relative to heterospecific pollen), should increase for plants that share many pollination partners.
<!-- DBS: Should we interchange amount and quantity so loosely like in the start of the sentence above? Also, it seems weird to call quantity of conspecific pollen a proxy for competition since it should be the inverse... -->
However, we also hypothesize that other factors like the plant's functional originality, its relative abundance, and the visit effectiveness should have the potential to compensate for this increase in competition. 
Third and finally, we examine how much the community context influences the way plants exploit these factors to minimise competition. 
Although some plant species might be widespread across multiple locations, competition for pollinators occurs at the community level and hence we predict the competition for pollination niche to be influenced by the local context.
<!-- DBS: I find the phrase "competition for pollination niche" to be very clunky. Can we do better? -->
If plant strategy is not flexible across the communities in our study\R{ref-plant-flexibility}, we could expect niches to be more similar within species than across species. 
On the contrary, if a species' strategy is flexible, we should expect plants to be able to occupy different niches in each community such that competition could ultimately be minimised regardless of the context. 
<!-- DBS: Why do we always say could and not would? -->

@@

We collected data from eleven co-flowering plant communities and their pollinators in three locations in the Argentinian Pampas. 
These locations are distributed across a precipitation gradient with the humid temperate climate of the eastern Flooding Pampas becoming drier as one moves inland and westward towards the Pampean Grasslands. 
This precipitation gradient generated a natural gradient of network architecture which allowed us to examine the generality of our findings.
<!-- DBS: "Natural gradient" of network architecture seems purposefully vague. -->
In each location, we sampled two restored and two agricultural fragments, except in the Flooding Pampas, where we were only able to sample one restored fragment due to the lack of available sites.

@@

Our first objective was to identify whether the net outcome of plants sharing pollinators at the community level is facilitation or competition. 
We use two complementary lines of evidence.  
The first can be found via the relationship between heterospecific and conspecific pollen deposited in plant stigmas [@tur_evaluating_2016]. 
This is because heterospecific pollen can be seen as an unavoidable cost that plants must pay for receiving conspecific pollen from shared pollinators. 
A negative relationship (when shared pollinators increase heterospecific pollen loads and decrease conspecific loads) may indicate that competition for pollination is strong.
In contrast, a positive relationship is what would be observed when facilitation is the dominant outcome.

@@

In each of the studied communities, we quantified pollen deposition in a subset of plant species between December 2010 and February 2011. 
This subset comprised between three and nine common insect-pollinated (entomophilous) plant species that were flowering during the sampling period (see Figure \@ref(fig:fig-trade-off-predictions)a). 
We chose plant species such that they cover a wide range on a specialization-generalization gradient as well as a wide range of abundances. 
In each of the selected plants, we removed all flowers leaving only buds that were expected to go into florescence on the next day. 
Approximately a quarter of these buds were bagged to prevent animal pollination. 
Two days after inflorescence, we collected all remaining flowers and counted the number of conspecific and heterospecific pollen grains in their pistils. 
More details can be found in @marrero_agricultural_2016.

@@

To evaluate the relationship between heterospecific and conspecific pollen, we used a generalised linear mixed model (GLMM) with a Poisson error distribution [@tur_evaluating_2016]. 
In this model, the number of conspecific pollen grains per flower was the response variable and the number of heterospecific grains the predictor variable (the number of stigmas sampled was added as an offset term). 
<!-- DBS: I have a technical question about this offset. Is it the raw count or the log of said count? -->
The slope and intercept of this relationship were allowed to differ between plant species within each community, which were modelled as a random effect. 
<!-- DBS: This is slightly ambiguous. Are communities or species the random effect? -->
Models were fitted using the function `glmer` from the R package `lme4 1.1-19` [@bates_fitting_2015]. 
A positive slope for a plant-community combination might indicate facilitation while a negative slope might indicate competition [@tur_evaluating_2016]. 
<!-- DBS: Everywhere else, we use definitive language about this interpretation. Here we temper it with 'might'. Why? -->

To compare the conspecific pollen deposited with and without animal-mediated pollination, we followed the same approach as above. 
In this case, however, the predictor variable was the treatment (whether the flower was bagged or unbagged) instead of the number of heterospecific grains.

@@

Our second objective was to investigate the relative contribution of different factors---that describe the plants' competition for pollination niches---to the pollination service. 
Generally speaking,
<!-- in the context of competition for pollination, -->
we expect that any factor that increases the amount of conspecific pollen deposited in stigmas, both in quantity and purity relative to heterospecific pollen, also has a positive effect on the pollination service.
<!-- DBS: There is no use to calling it something succinct like purity if we always have to say "purity relative to heterospecific pollen"...  -->
Specifically, we investigated the effect of (*i*) a plant's number of shared pollinator species, (*ii*) a plant's abundance relative to the rest of the community, (*iii*) the mean visit effectiveness---a metric that combines
the share of pollen that a plant species is able to place on each of their floral visitors
<!-- DBS: The line above is very confusing and took me about five reads before I understood what you wanted to say. We need to revise it. -->
and the number of visits it receives from them, and (*iv*) the plant's functional originality [@laliberte_distance-based_2010]. 
See *Data Analysis* section below for more details on these four variables. 
<!-- Here, we define visit effectiveness of the pollinator species *i* to plant species *j* as the proportion between the visits of *i* to *j* and all visits of *i*, multiplied by the proportion between the number of pollen grains of *j* attached to individuals of *i* and all pollen attached to individuals of *i*. -->

@@

To obtain the number of shared pollinators for each species, we collected data to construct qualitative and quantitative pollination networks. 
Qualitative networks were constructed based on two-hour observations of floral visits in each fragment. 
Quantitative networks were constructed using two 50 m randomly located transects in each fragment. 
We counted and collected all floral visitors found in a 2 m wide strip while walking at a pace of 10 m per minute [@marrero_effect_2014; @memmott_structure_1999]. 
We visited the transects each month between November 2010 and March 2011. 
To obtain plant abundance, we counted all units of floral attraction found during an independent sampling of the same transects used to construct the quantitative visitation networks. 
To estimate visit effectiveness, we need to construct pollen transfer networks in addition to the visitation networks. 
To do this, we examined the pollen loads present on the floral visitors collected [@marrero_exotic_2017]. 
When the pollen count on an individual animal was estimated to be less than 2,000 grains, we identified every grain to the species level when possible and to pollen complexes when it was not. 
When the pollen count was above 2,000 grains, we classified approximately 50% of pollen and total pollen counts were extrapolated [@bosch_plant-pollinator_2009]. 
Finally, we also recorded morphological traits that relate to plant type (herb, shrub, climber), life cycle (annual, perennial), flower colouration, phenology, and whether the species is native in the study region\R{justification-traits}. 
<!-- We used those traits because they were the only ones available. <!--.. (TODO: Add a better justification for the traits used)--> 
<!-- DBS: I bet we can get away with submitting without justifying the traits. -->
More details can be found in Marrero *et al.* (2014 and 2017).

@@

To investigate the impact of ecological factors on pollination service, we used two sets of linear mixed models (LMM) with bootstrap resampling.
The response variables for these model sets were the number of conspecific and heterospecific pollen grains deposited per stigma in flowers open to animal-mediated pollination. 
We used LMMs in which pollen loads were log-transformed because these models offered a better fit than equivalent GLMMs with Poisson (or quasi-Poisson) error structure. 
Models were fitted using the R package `nlme 3.1-131` [@pinheiro_nlme_2018].
